---
title: "Instructions for installing keras3"
author: "Maria Anastasiadi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To be able to complete the practical on Artificial Neural Networks you will need to download and install python and then create a virtual environment to run python. 
First make sure you have git installed on your computer. If not go to https://git-scm.com/download/win for installing git on windows or here https://git-scm.com/download/mac for macOS. 

Next install and load the library `reticulate`
```{r}
if(!require('reticulate')) install.packages('reticulate'); library('reticulate')

install_python("3.11:latest") # install latest patch available at python.org

# select the latest 3.11.* patch installed locally
virtualenv_create("r-reticulate", version = "3.11")
use_virtualenv("r-reticulate")
```

The next step is to install `keras3` and `tensorflow`. 

```{r}
install.packages("keras3")
```

```{r}
keras3::install_keras(backend = "tensorflow")
```

Load keras3 
```{r}
library(keras3)  
```

Now you should be ready to run keras3 in your computer. 

