---
title: "01_objective_storage_ripe_classification_models"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Use the apple physicochemical data from the DCA trial Year 1 and Year 2.


```{r}
# load libraries
library(ggplot2)
library(caret)
library(dotenv)
library(dplyr)
```

################################################
################### Load .env ##################
################################################

```{r}
cat("Loading environment variables from .env file...\n")
tryCatch(
  {
    dotenv::load_dot_env()
    cat("Environment variables loaded successfully.\n")
  },
  error = function(e) {
    cat("Error loading environment variables:", e$message, "\n")
    cat("Loading from hardcoded path as fallback...\n")
    dotenv::load_dot_env("/Users/mspriggs/Library/CloudStorage/OneDrive-Illumina,Inc./Documents/Applied_Bioinformatics/modules/machine_learning_for_metabolomics/assignment/scripts/.env") # nolint
    cat("Loaded .env from hardcoded path\n")
  }
)
```

################################################
############# Set up Global Scope ##############
################################################

```{r}
cat("Setting up global variables and sourcing utility functions...\n")
WORKING_DIR <- Sys.getenv("WORKING_DIR")

# define file paths
UTILS_FILEPATH <- file.path(WORKING_DIR, "scripts", "00_data_prep_utils.R")
DATA_DIR <- file.path(WORKING_DIR, "Assignment_Data")
OUTPUT_DIR <- file.path(WORKING_DIR, "outputs")
PLOTS_DIR <- file.path(OUTPUT_DIR, "plots")

date_str <- format(Sys.Date(), "%Y-%m-%d")
```

You will use the data from Year1 to train and optimise the classification model and the data from Year2 to validate its performance

```{r}
# source utility functions
source(UTILS_FILEPATH)
print(paste("Utility functions sourced from:", UTILS_FILEPATH))
```

################################################
################### Main #######################
################################################


```{r}
# load DCA Year one as train data
training_data <- load_data(file.path(DATA_DIR, "DCA_Year1.csv"))

# load DCA Year two as test data
test_data <- load_data(file.path(DATA_DIR, "DCA_Year2.csv"))
```

```{r}
# remove missing values from DCA Year 1
training_data <- training_data %>% drop_na()
```

```{r}
# infer mean for na values in DCA Year 2
test_data <- infer_mean_for_nan(test_data)
```



```{r}
# Fix column name mismatch
colnames(training_data)[colnames(training_data) == "X7OH.ABA"] <- "X7.OH.ABA"
```

```{r}
# extract features and labels
training_samples <- extract_sample_names(training_data)
rownames(training_data) <- training_samples
training_data <- training_data[, -1]

test_samples <- extract_sample_names(test_data)
rownames(test_data) <- test_samples
test_data <- test_data[, -1]
```

```{r}
# convert storage.week and ripening.stage to factors
training_data <- mutate_df_factors(training_data)
test_data <- mutate_df_factors(test_data)
```

```{r}
# Check the proportion of each class in the dataset for training_data
round(prop.table(table(training_data$Storage.Week)) * 100, 2)
```

```{r}
# get number of samples in each class
table(training_data$Storage.Week)
```

good balance. Week 0 has a slightly higher proportion, but this isn't a major concern.

```{r}
round(prop.table(table(training_data$Ripening.Stage)) * 100, 2)
```

```{r}
# get number of samples in each class
table(training_data$Ripening.Stage)
```

early stage is overrepresented, focus on storage week classification first

################################################
################ Data Preparation ##############
################################################

First model: classify storage week using Storage.Week as the label and all other features as predictors
```{r}
# remove ripening stage, use only storage week for classification
training_features <- training_data %>%
  dplyr::select(-Ripening.Stage)
```



###############################################################
################ Model Training (Storage Week) ################
###############################################################

```{r}
# rename Storage.Week levels to valid R variable names
training_features$Storage.Week <- factor(
  training_features$Storage.Week,
  levels = levels(training_features$Storage.Week),
  labels = paste0("Week_", levels(training_features$Storage.Week))
)
```


```{r}
# use the time splits for cross-validation
train_control_time <- trainControl(
  method = "cv",
  initialWindow = floor(0.8 * nrow(training_features)),
  horizon = floor(0.2 * nrow(training_features)),
  fixedWindow = FALSE,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final"
)

set.seed(123) # for reproducibility
# Define parameter grid
tune_grid <- expand.grid(mtry = c(5, 10, 15, 20, 25))

# Train with grid search
rf_model <- train(
  Storage.Week ~ .,
  data = training_features,
  method = "rf",
  trControl = train_control_time,
  tuneGrid = tune_grid,
  ntree = 500
)
```

################################################
########## Test Model with DCA Year 2 ##########
################################################

```{r}
# Prepare test data
test_features <- test_data %>%
  dplyr::select(-Ripening.Stage)
```


```{r}
# rename Storage.Week levels to match training data
test_features$Storage.Week <- factor(
  test_features$Storage.Week,
  levels = levels(test_features$Storage.Week),
  labels = paste0("Week_", levels(test_features$Storage.Week))
)
```

```{r}
# Predict on test data
test_predictions <- predict(rf_model, newdata = test_features)
test_probabilities <- predict(rf_model, newdata = test_features, type = "prob")
```

```{r}
# Evaluate performance
conf_matrix <- confusionMatrix(test_predictions, test_features$Storage.Week)
print(conf_matrix)
```

#################################################################
################ Model Training (Ripening Stage) ################
#################################################################

```{r}
# remove storage week, use only ripening stage for classification
training_features_ripening <- training_data %>%
  dplyr::select(-Storage.Week)
```

```{r}
train_control <- trainControl(
  method = "cv",
  initialWindow = floor(0.8 * nrow(training_features_ripening)),
  horizon = floor(0.2 * nrow(training_features_ripening)),
  fixedWindow = FALSE,
  classProbs = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final"
)

# Calculate class weights
class_counts <- table(training_features_ripening$Ripening.Stage)
class_weights <- max(class_counts) / class_counts

set.seed(123) # for reproducibility
# Define parameter grid
tune_grid_ripening <- expand.grid(mtry = c(5, 10, 15, 20, 25))

# Train with grid search
# rf_model_ripening <- train(
#   Ripening.Stage ~ .,
#   data = training_features_ripening,
#   method = "rf",
#   trControl = train_control,
#   tuneGrid = tune_grid_ripening,
#   ntree = 500,
#   weights = class_weights
# )

# Train with class weights
rf_model_final <- randomForest(
  Ripening.Stage ~ .,
  data = training_features_ripening,
  ntree = 500,
  mtry = 20,
  classwt = class_weights,
  importance = TRUE
)
```

#################################################
########## Test Ripening Stage Model with DCA Year 2 ##########
#################################################

```{r}
# Prepare test data
test_features_ripening <- test_data %>%
  dplyr::select(-Storage.Week)
```

```{r}
# predict on test data
test_predictions_ripening <- predict(rf_model_ripening, newdata = test_features_ripening)
test_probabilities_ripening <- predict(rf_model_ripening, newdata = test_features_ripening, type = "prob")
```

```{r}
# evaluate performance
conf_matrix_ripening <- confusionMatrix(test_predictions_ripening, test_features_ripening$Ripening.Stage)
print(conf_matrix_ripening)
```